The markings of the trade: exploring the patterns of olive oil production in Roman Baetica

### 2018-06-22 Paper/jaccardbaetica.tex
add new changes

Roman province such as Baetica (currently Andalusia, south Spain) became important production centers of olive oil during the Roman Empire. Olive oil was considered as the liquid gold since it was used in different aspect of the daily life as cooking, hygiene or lighting  \citep{mattingly_d.j._oil_1988}. As a consequence, the high demand of Roman provinces stimulated by the good condition of the Baetica lands allowed to develop a massive infrastructure of olive oil production. This product was distributed in large amount of amphorae along the province, mostly to supply the Roman Army and Italy \citep{blazquez_exportacion_1980}.

The production and distribution of olive oil in this ancient province were growing exponentially during almost three centuries \citep{remesal_concierto}. As a result, hundred of amphora workshops were made to support this high demand of Roman Empire. These workshops were located along the rivers Guadalquivir and Genil, supplying from the riverine connectivity to the Mediterranean and Atlantic routes \citep{garcia_vargas_enrique_formal_2010}.

The majority of olive oil amphorae produced in this province and shipped through the provinces belong to Dressel 20 type \citep{dressel,martin-kilcher_romischen_1994}. This amphora is commonly associated with the transportation of Baetican olive oil during the Roman Empire \citep{berni_millet_epigrafianforica_2008}. Most Dressel 20 were marked in stamps, inked in \textit{tituli picti} and incise in \textit{graffiti} with different information. However, inscriptions and \textit{tituli picti} have not enough analysed by researchers due to the fragmentation of the material or the shortage of samples \citep{aguilera_evolucion_2007,rovira_guardiola_grafitos_2007}. By contrast, stamps are the most studied in this type of amphorae. During almost three centuries, stamps were used to mark amphorae with a different chronological frequency \citep{remesal_sellar_2016}. Frequently, they were marked mainly in handles but rarely in rims and body \citep{millet_anforas_1998}. 
The information of the stamps is shown in different forms and letter content and it seems that there was not a unique criterion. Stamps was mostly formed by a code of three letters and they can appear in a abbreviated form or complete and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}.

%Many hypothesis involved the meaning of stamps. 
 There is not a general consensus about the meaning of the stamps \citep{rodriguez_baetican_1998}. On the one hand, stamps are identified as the land owner of the olive groves \citep{rodriguez_economioleicola_1977}. On the other, they could be the owner of the making amphorae workshops or even a production counting system  \citep{berni_millet_epigrafianforica_2008}. In any case, the use of these stamps became in a good proxy to define somehow the system of working in the workshops.

%Hyphotesis
Nevertheless, some challenges remain under discussion such as how this province was organized and whether it is possible to distinguish production patterns in the olive oil trade. Our questions will be focused on the distribution of amphoric stamps. Did they follow some distribution pattern? Did stamps share the same workshop? Neither written sources have been found that it could explain the economic role of Baetica province in the Roman organization. On the other hand, archaeological evidence shows a highly specialized production with a long activity in a this specific area with apparently few changes \citep{remesal_anforas_2004}.

%hipotesis: si esto es así entonces es asá

Here, therefore, this study proposes an evolutionary approach to explore the effect of Baetican olive oil production by computing of dissimilarity index. The assumption of this analysis is that closer workshops concentrate similar amphoric stamps in a specific area than the farthest workshops. By contrast, if we found similar amphoric stamps in different areas between workshops then the correlation between spatial distance would not be valued.

@@

The chronology in the workshops is widely diverse from the first to the third centuries AD \citep{millet_anforas_1998,rodriguez_baetican_1998,chic2005comercio}. Some stamps show a more specific chronology while the majority of them display a large activity of production being difficult to specify an accurate chronology. This could be due to two reasons. Firstly, most of the workshops were partially excavated and focused on archaeological surveys in order to collect the maximum stamps as possible. Secondly, this amphora type was produced during almost three centuries with apparently few changes so they presented a long chronology. (podría citar nuestro paper por toda la face).

We studied a dataset of 3787 stamps collected from different Dressel 20 amphora workshops in Baetica province. The stamp database was compiled by CEIPAC database \citep{remesal_centro_2015} (see CEIPAC database here \url{http://romanopendata.eu}). However, approximately the 70 \% of stamps cannot be tested due to fragmentation or incomplete information. Consequently, we discard integrate the fragmented stamps in our dataset. We finally filter a total sample of 987 stamps composed by 131 different stamps from 81 workshops.

From the database, we collected the site where it was found and the stamp code. We also created a new column with the area where stamps were found. These areas, known as \textit{conventus}, were administrative centres for territorial organization in the Roman Empire. Dressel 20 stamps were found in three different \textit{conventus}: \textit{Hispalensis} (currently Seville, hereafter Hispalis), \textit{Cordubensis} (currently C\'ordoba, hereafter Corduba) and \textit{Astigi} (currently Écija, Sevilla, hereafter Astigi) \citep{rodriguez_economioleicola_1977,chic_datos_2001,berni_millet_epigrafianforica_2008} .

The distribution of amphora stamps in different \textit{conventus} can be seen in Fig. \ref{frequency}. The majority of stamps found are concentrated in \textit{Hispalis} with 574 stamps while \textit{Corduba} and \textit{Astigi} with 267 and 146 stamps, respectively. Mostly workshops show a homogeneity on frequency of stamps except La Catria and Arva (both show 29 different stamps) that show a big amount of different stamps. According to some researchers, those workshops became in most important centres of amphora production although they could have been more intensely prospected then others. \citep{arva_1997}.

@@

Considering our dataset as non-uniform sample, this method provides a useful tool to handle large samples with different sizes and diversity \citep{wolda_similarity_1981}. Morisita-Horn index can be expressed considering 0 as total presence of similarity of stamps and 1 a totally dissimilarity between stamps. In our case, it will be calculated the number of times that one stamp appear in a amphora workshop. This method allows to bear in mind the similar number of times for each repeated stamp per workshop. If two workshops have similar stamp codes then the probability would be 0 whereas stamps codes are totally different when the results would be 1.

@@

The analysis shows that amphoric stamps could be correlated with the spatial distance. The correlation coefficients range from a minimum to a maximum. The dendrogram shown in Fig. \ref{dendro} was obtained with Morisita-Horn index. The dendrogram suggests that amphora workshops used different stamps for their production system. Nearby workshops show a similarity on the stamps while most of them seem to display different stamps roles. Additionally, no groups of similar stamps were found in the cluster. The majority of stamp grouping was composed by no more than three workshops. Indeed workshops that shared more similar amphoric stamps belonged to the same \textit{conventus} area, such as Picachos, Cerro de los Pesebres and El Castillejo.

@@

In our analysis, no strong relation between stamps and spatial distance were detected in amphora workshops. The analysis suggests that there is not connection between stamps and the same amphora workshops, excluding certain exceptions when nearby workshops share the same amphoric stamp. Consequently, the majority of stamps are located in different amphora workshops and only similar stamps between closer amphora workshops were found. In any case, our results show that most similar stamps were detected in the same \textit{conventus} area. These stamps tend to share the same area of production but there is not a general relation between groups of amphora workshops and area.

@@

The results of the case study could be interpreted due to several reason according to previous hypotheses. On the one hand, the use of these amphoric stamps could have been exclusively running by the owner or owners of the workshop to distinguish the amphora workshop (CITAR). This hypothesis would explain the fact that we do not find similar stamps in different workshops, however, we found different amphora stamps in the same workshops that they would be barely difficult to assign different owners. Neither a quality distinction does seem to be used to specify the value of the product \citep{callender}. On the other hand, it could be interpreted somehow a batch systematic organization. Considering that Dressel 20 was not marked in most cases, some authors point out that potters marked amphorae to prepare and distribute the commodity to be shipped \citep{berni_millet_epigrafianforica_2008}. This method would be used as an identifier to count the number of amphorae of a branch (CITAR). This method could also have served to identify different groups of potter workers working in the same amphora workshop. Potters could have marked the amphorae to distinguish different groups working in parallel \citep{li_crossbows_2014}. This it could be explain wherefore we found different stamps in a same workshop. 
In any case, we do not have enough archaeological evidence that can validate the interpretations presented here and our results are certainly valid only with the context of our case study.

As summary, this method presented here provides a potential tool to understand mechanisms of production based on the similarity of artefacts. This method have identified differences in the case of the amphoric production within Roman Empire. Accordingly the results have highlighted for the interpretation of the complex economic processes connected with the archaeological evidence.

>>>>

Roman provinces such as Baetica (currently Andalusia, south Spain) became important production centres of olive oil during the Roman Empire. Olive oil was considered as the liquid gold since it was used in different aspect of the daily life as cooking, hygiene or lighting  \citep{mattingly_d.j._oil_1988}. As a consequence, the high demand of Roman provinces stimulated by the good condition of the Baetica lands allowed to develop a massive infrastructure of olive oil production. This product was distributed in large amount of amphorae along the province, mostly to supply the Roman Army and Italy \citep{blazquez_exportacion_1980}.

The production and distribution of olive oil in this ancient province were growing exponentially during almost three centuries \citep{remesal_concierto}. As a result, thousands of amphora workshops were made to support this high demand of Roman Empire. These workshops were located along the rivers Guadalquivir and Genil, supplying from the riverine connectivity to the Mediterranean and Atlantic routes \citep{garcia_vargas_enrique_formal_2010}.

The majority of olive oil amphorae produced in this province and shipped through the provinces belong to Dressel 20 type \citep{dressel,martin-kilcher_romischen_1994}. This amphora is commonly associated with the transportation of Baetican olive oil during the Roman Empire \citep{berni_millet_epigrafianforica_2008}. Most Dressel 20 were marked in stamps, inked in \textit{tituli picti} and incise in \textit{graffiti} with different information. However, inscriptions and \textit{tituli picti} have not enough analysed due to the fragmentation of the material or the shortage of samples \citep{aguilera_evolucion_2007,rovira_guardiola_grafitos_2007}. By contrast, stamps are the most studied in this type of amphorae. During almost three centuries, stamps were used to mark amphorae with a different chronological frequency \citep{remesal_sellar_2016}. Frequently, they were marked mainly in handles but seldom in rims and body \citep{millet_anforas_1998}. 
The information of the stamps is shown in different forms and letter content and it seems that there was not a unique criterion. Stamps are mostly formed by a code of three letters and they can appear in a abbreviated form or complete and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}.

%Many hypothesis involved the meaning of stamps. 
 There is not a general consensus about the meaning of the stamps \citep{rodriguez_baetican_1998}. It seems that stamps could have been identified as the land owner of the olive groves \citep{rodriguez_economioleicola_1977}. On the other hand, they could belong to the owner of the making amphorae workshops or even a production counting system  \citep{berni_millet_epigrafianforica_2008}. In any case, the use of these stamps became in a good proxy to define somehow the system of working in the workshops.

%Hyphotesis
Nevertheless, some challenges remain under discussion such as how this province was organized and whether it is possible to distinguish production patterns in the olive oil trade. Our questions will be focused on the distribution of amphoric stamps. Did they follow a distribution pattern? Did stamps share the same workshop? Neither written records have been found that it could explain the economic role of Baetica province in the Roman organization. Indeed, archaeological evidence shows a highly specialized production with a long activity in a this specific area with apparently few changes \citep{remesal_anforas_2004}.

%hipotesis: si esto es así entonces es asá
Here, therefore, this study proposes an evolutionary approach to explore the effect of Baetican olive oil production by computing of dissimilarity index. The assumption of this analysis is that closer workshops concentrate similar amphoric stamps in a specific area than the distant workshops. By contrast, if we found similar amphoric stamps in different areas between workshops then the correlation between spatial distance would not be valued.

@@

The chronology in the workshops is widely diverse from the first to the third centuries AD \citep{millet_anforas_1998,rodriguez_baetican_1998,chic2005comercio}. Some stamps show a more specific chronology while the majority of them display a large activity of production that it can be difficult for specifying an accurate chronology. This could be due to two reasons; firstly, most of the workshops were partially excavated and focused on archaeological surveys in order to collect the maximum stamps as possible; secondly, Dressel 20 was produced during almost three centuries with apparently few changes. (podría citar nuestro paper por toda la face).

We studied a dataset of 3787 stamps collected from different Dressel 20 amphora workshops in Baetica province. The stamp database was compiled by CEIPAC database \citep{remesal_centro_2015} (see CEIPAC database here \url{http://romanopendata.eu}). However, approximately the 70 \% of stamps cannot be tested due to fragmentation or incomplete information. Consequently, we discard integrate the fragmented stamps in our dataset. We finally filter a total sample of 987 stamps comprised of 131 different stamps from 81 workshops.

From the database, we collected the site where stamps were found and the stamp code. We also created a new column with the area of the stamps. These areas, known as \textit{conventus}, were administrative centres for territorial organization in the Roman Empire. Dressel 20 stamps were found in three different \textit{conventus}: \textit{Hispalensis} (currently Seville, hereafter Hispalis), \textit{Cordubensis} (currently C\'ordoba, hereafter Corduba) and \textit{Astigi} (currently Écija, Sevilla, hereafter Astigi) \citep{rodriguez_economioleicola_1977,chic_datos_2001,berni_millet_epigrafianforica_2008} .

The distribution of amphora stamps in different \textit{conventus} can be seen in Fig. \ref{frequency}. The majority of stamps found are concentrated in \textit{Hispalis} with 574 stamps while \textit{Corduba} and \textit{Astigi} with 267 and 146 stamps, respectively. Mostly, workshops show a homogeneity on the frequency of stamps except both La Catria and Arva that show a big amount of stamps with 29 different stamps. According to previous studies, those workshops became in most important centres of amphora production although they could have been more intensely prospected then others. \citep{arva_1997}.

@@

Considering our dataset as a non-uniform sample, this method provides a useful tool to handle large samples with different sizes and diversity \citep{wolda_similarity_1981}. Morisita-Horn index can be expressed considering 0 as total presence of similarity of stamps and 1 a totally dissimilarity between stamps. In our case, it will be calculated the number of times that one stamp appear in an amphora workshop. This method allows to bear in mind the similar number of times for each repeated stamp per workshop. If two workshops have similar stamp codes then the probability would be 0 whereas stamps codes are totally different when the results would be 1.

@@

The analysis shows that amphoric stamps could be correlated with the spatial distance. The correlation coefficients range from a minimum to a maximum. The dendrogram shown in Fig. \ref{dendro} was obtained with Morisita-Horn index. The dendrogram suggests that amphora workshops used different stamps for their production system. Nearby workshops show a similarity on the stamps while most of them seem to display different stamps roles. Additionally, groupings of similar stamps were not found in the cluster. The majority of stamp grouping was composed by no more than three workshops. Indeed workshops that shared more similar amphoric stamps belonged to the same \textit{conventus} area, such as Picachos, Cerro de los Pesebres and El Castillejo.

@@

In our analysis, no strong relation between stamps and spatial distance have been detected in amphora workshops. The analysis suggests that there is not connection between stamps and the same amphora workshops, excluding certain exceptions when nearby workshops share the same amphoric stamp. Consequently, the majority of stamps are located in different amphora workshops and only similar stamps between closer amphora workshops were found. In any case, our results show that most similar stamps were detected in the same \textit{conventus} area. These stamps tend to share the same area of production but there is not a general relation between groups of amphora workshops and area.

@@

The results of the case study could be interpreted by several reasons. On the one hand, the use of these amphoric stamps could have been exclusively running by the owner or owners of the workshop to distinguish the amphora workshop (CITAR). This hypothesis would explain the fact that we do not find similar stamps in different workshops, however, we found different amphora stamps in the same workshops that they would be barely difficult to assign different owners. Neither a quality distinction does seem to be used to specify the value of the product \citep{callender}. On the other hand, it could be interpreted somehow a batch systematic organization. Considering that Dressel 20 was not marked in most cases, some authors point out that potters marked amphorae to prepare and distribute the commodity to be shipped \citep{berni_millet_epigrafianforica_2008}. This method would be used as an identifier to count the number of amphorae of a branch (CITAR). This method could also have served to identify different groups of potter workers working in the same amphora workshop. Potters could have marked the amphorae to distinguish different groups working in parallel \citep{li_crossbows_2014}. This it would explain wherefore why we detect different stamps in a same workshop. 
In any case, we do not have enough archaeological evidence that can validate the interpretations presented here and our results are certainly valid only with the context of our case study.

As summary, this method presented here provides a potential tool to understand mechanisms of production based on the similarity of artefacts. This method have identified differences in the case of the amphoric production within Roman Empire. Accordingly, the results have highlighted for the interpretation of the complex economic processes connected with the archaeological evidence.
